Statistical analysis of periodic data in neuroscience

### 2020-12-31 manuscript/manuscript.Rmd
Fixed cluster correction function

---
title: "Statistical analysis of periodic data in neuroscience"
author: "Daniel H. Baker"
date: "13/12/2020"
output:
  bookdown::pdf_document2:
    fig_caption: yes 
    toc: false   
    keep_tex: true
  html_document:
    df_print: paged
  pdf_document: default
bibliography: bibliography.bib
biblio-style: apalike
---

@@

packagelist <- c('bookdown','pals','shape','diagram','HDMD','FourierStats')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

v4Interp <- function(df, xo, yo, rmax = .75, gridRes = 67) {
  ## Create a function to perform Matlab's v4 interpolation.
  ## Takes as input a data-frame with columns x, y, and z (x co-ordinates, y co-ordinates, and amplitude)
  ## and variables xo and yo, the co-ordinates which will be use to create a grid for interpolation
  xo <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
  yo <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
  xy <- df$x + df$y*sqrt(as.complex(-1))
  d <- matrix(rep(xy,length(xy)),nrow = length(xy), ncol = length(xy))
  d <- abs(d - t(d))
  diag(d) <- 1
  g <- (d^2) * (log(d)-1)   # Green's function.
  diag(g) <- 0
  weights <- qr.solve(g,df$z)
  xy <- t(xy)
  outmat <- matrix(nrow = gridRes,ncol = gridRes)
  for (i in 1:gridRes){
    for (j in 1:gridRes) {
      test4 <- abs((xo[i,j] + sqrt(as.complex(-1))*yo[i,j]) - xy)
      g <- (test4^2) * (log(test4)-1)
      outmat[i,j] <- g %*% weights}}
  outDf <- data.frame(x = xo[,1],outmat)
  names(outDf)[1:length(yo[1,])+1] <- yo[1,]
  return(outDf)}

@@

More recently, Victor and Mast [-@Victor1991] proposed a simpler version of $T^2$, called $T^2_{circ}$. The $T^2_{circ}$ statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance. When these conditions are met, the test statistic is defined as:

@@

where $x_j$ denotes the $j$th observation of the dependent variables, and all other terms retain their previous meanings. Notice that no covariance term is present in equation \@ref(eq:t2c), because of the independence assumption. This makes the statistic simpler to calculate, but causes problems when the assumption is violated.

@@

All scripts used to generate this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats). These include _R_ functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.

@@

The upper row of Figure \@ref(fig:amphists) shows scatterplots of simulated Fourier components, expressed using real (x) and imaginary (y) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \@ref(fig:amphists) shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d > 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]). One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.

@@

Here we replicate and extend these simulations (see Figure \@ref(fig:powerfig)), and confirm that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f). However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N > 32) and large effect sizes (d > 2 when N > 8). These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for sample sizes, so its use should be restricted to small sample studies.

@@

Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on meeting the test's more stringent assumptions. The two variables must be independent, and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \@ref(fig:falsealarms) shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (dashed line). However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the variances of the two dependent variables more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2. In contrast, the $T^2$ statistic, which explicitly takes account of the covariance matrix (see equation \@ref(eq:t2eq)) shows no increase (black curves).

```{r falsealarms, fig.width=6, fig.height=9, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b). Estimates are for 100000 simulations per condition, with N=10 observations. The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}

@@

  A <- cov(compdata)
  ctr    <- colMeans(compdata) 
  RR     <- chol(A)                               # Cholesky decomposition
  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
  eigVal  <- eigen(A)$values
  eigVec  <- eigen(A)$vectors
  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
  ellRot  <- eigVec %*% t(ellBase)

  points(0.05*xdata[1:100]+corvals[cond],0.005*ydata[1:100]+0.02,pch=16,cex=0.5,col=nbdtpal[3])
  matlines(0.1*xMat+corvals[cond], 0.01*yMat+0.02, lty=1, lwd=3, col="black")
  lines(0.1*(ellRot+ctr)[1, ]+corvals[cond], 0.01*(ellRot+ctr)[2, ]+0.02, lwd=2)
}
text(-0.95,0.095,'(a)',cex=1.5)

@@

  A <- cov(compdata)
  ctr    <- colMeans(compdata) 
  RR     <- chol(A)                               # Cholesky decomposition
  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
  eigVal  <- eigen(A)$values
  eigVec  <- eigen(A)$vectors
  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
  ellRot  <- eigVec %*% t(ellBase)

  points(0.05*xdata[1:100]+corvals[cond],0.005*ydata[1:100]+0.02,pch=16,cex=0.5,col=nbdtpal[3])
  matlines(0.1*xMat+corvals[cond], 0.01*yMat+0.02, lty=1, lwd=3, col="black")
  lines(0.1*(ellRot+ctr)[1, ]+corvals[cond], 0.01*(ellRot+ctr)[2, ]+0.02, lwd=2)
}
text(-0.95,0.095,'(b)',cex=1.5)

@@

plotlims <- c(0,8,0,2) 
ticklocsx <- seq(0,8,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Outlier Mahalanobis distance (D)", col.lab=rgb(0,0,0), line=1.2, cex.lab=1.5)
mtext(text = "Proportion significant", side = 2, at=0.5, line=1.5, cex=1.5)
points(0,2,pch=15,col=rgb(0.99,0.99,0.99),cex=0.1)

@@

  polygon(0.4*(ellRot2+ctr2)[1, ]+xlocs[m], 0.1*(ellRot2+ctr2)[2, ]+1.5, col=addalpha(nbdtpal[2],0.3),border=NA)
  lines(0.4*(ellRot+ctr)[1, ]+xlocs[m], 0.1*(ellRot+ctr)[2, ]+1.5, lwd=1)
  lines(0.4*c(ctr2[1],xdata[1])+xlocs[m],0.1*c(ctr2[2],ydata[1])+1.5,col=nbdtpal[2],lwd=3)
  points(0.4*xdata+xlocs[m],0.1*ydata+1.5,pch=16,cex=0.5,col=nbdtpal[3])
  points(0.4*ctr2[1]+xlocs[m],0.1*ctr2[2]+1.5,pch=16,cex=1,col=nbdtpal[1])
  points(0.4*xdata[1]+xlocs[m],0.1*ydata[1]+1.5,pch=16,cex=1,col=nbdtpal[2])
}

  text(1,1.8,'D = 1',cex=1.5,adj=0.5)
  text(3.5,1.8,'D = 3',cex=1.5,adj=0.5)
  text(6,1.8,'D = 5',cex=1.5,adj=0.5)

@@

@@

Following ${ANOVA}^2_{circ}$, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that standard Bonferroni corrections are applied. Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary). A repeated measures version of ${ANOVA}^2_{circ}$ has also been implemented following the same logic. The project repository contains _R_ scripts for both of these functions.

@@

Figure \@ref(fig:mousedata)a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz. There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality. Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = 0.84, _df_ = 5, _p_ = 0.44). However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \@ref(fig:mousedata)b). The condition index test was non-significant for both conditions (sound: CI = 1.59, _p_ = 0.66; light: CI = 1.69, _p_ = 0.59), so a paired-samples $T^2_{circ}$ test was conducted. This revealed a significant difference between conditions ($T^2_{circ}$ = 1.39, $F_{(2,10)}$ = 8.32, _p_ = 0.007) with an effect size of D = 1.02. This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al. [-@Hwang2019] went on to explore interactions between these two signals.

@@

load('Hwangdata.RData')

@@

ellRot1 <- getel(data.frame(xdata,ydata))
ellRot2 <- getel(data.frame(xdata2,ydata2))
polygon(10*(ellRot1[1, ]+mean(xdata))+120, 1+(ellRot1[2, ]+mean(ydata))/3, col=addalpha(nbdtpal[2],0.3),border=NA)
polygon(10*(ellRot2[1, ]+mean(xdata2))+120, 1+(ellRot2[2, ]+mean(ydata2))/3, col=addalpha(nbdtpal[3],0.3),border=NA)

@@

# totest <- data.frame(xdata,ydata)
# totest2 <- data.frame(xdata2,ydata2)
# CI.test(totest)
# CI.test(totest2)
# tsqc.test(totest,y=totest2,paired=TRUE)
# 
# abs1 <- abs(rowMeans(mousedata[,1:2,11,f1]))
# abs2 <- abs(rowMeans(mousedata[,1:2,12,f1]))
# t.test(abs1,abs2,paired=TRUE)

# colnames(totest2) <- colnames(totest)
# fores <- rbind(totest,totest2)
# mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=6))
# D <- sqrt(mahala_sq$distance[1,2])

@@

@@

Vilidaite et al. [-@Vilidaite2018] measured visual responses to flickering grating stimuli in a large sample of 100 adults. Each participant completed a series of 11-second trials, in which stimuli of different contrasts flickered at 7Hz (on-off sinusoidal flicker). Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \@ref(fig:humanSSVEP)), were well-isolated in the Fourier domain, and increased with stimulus contrast. For this analysis, we took responses from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), averaged across repetition for each participant. Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \@ref(fig:humanSSVEP). Any participant that contributed at least one outlier was excluded, leaving a total of 89 participants for the main analysis.

```{r humanSSVEP, fig.cap = 'Summary of human SSVEP data. Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes). Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point). Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.', fig.width=14, fig.height=5.4, fig.align="center", echo=FALSE}

@@

@@

plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,8), ylim=c(0,2))
nexcluded <- NULL
isincluded <- 1 + 0*(1:100)

for (cond in 1:7){
datatoplot <- abs(apply(alltarget[,1:64,cond,], 2, mean))
datatoplot[c(13,19)] <- mean(datatoplot)   # set the mastoids to 0
testDat<- data.frame(x = montage$electrodelocs[montage$channelmappings,1],
                     y = -montage$electrodelocs[montage$channelmappings,2],
                     z = datatoplot)

@@

points(cond+montage$electrodelocs[blackelectrodes,1]*0.8,1.5+montage$electrodelocs[blackelectrodes,2]*0.8,pch=16,col="black",cex=1.5)

@@

for (cond in 1:7){
  xvals <- Re(cleandata[,cond])/5
  yvals <- Im(cleandata[,cond])/5
  ellRot <- getel(data.frame(xvals,yvals))
  lines((ellRot[1, ]+mean(xvals))+cond, (ellRot[2, ]+mean(yvals))+0.5)

>>>>

---
title: "Statistical analysis of periodic data in neuroscience"
author: "Daniel H. Baker"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    fig_caption: yes 
    toc: false   
    keep_tex: true
  html_document:
    df_print: paged
  pdf_document: default
bibliography: bibliography.bib
biblio-style: apalike
---

@@

packagelist <- c('bookdown','pals','pracma','shape','reshape2','diagram','HDMD','FourierStats')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

v4Interp <- function(df, xo, yo, rmax = .75, gridRes = 67) {
  ## helper function to perform Matlab's v4 interpolation.
  ## Takes as input a data-frame with columns x, y, and z (x co-ordinates, y co-ordinates, and amplitude)
  ## and variables xo and yo, the co-ordinates which will be use to create a grid for interpolation
  xo <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
  yo <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
  xy <- df$x + df$y*sqrt(as.complex(-1))
  d <- matrix(rep(xy,length(xy)),nrow = length(xy), ncol = length(xy))
  d <- abs(d - t(d))
  diag(d) <- 1
  g <- (d^2) * (log(d)-1)   # Green's function.
  diag(g) <- 0
  weights <- qr.solve(g,df$z)
  xy <- t(xy)
  outmat <- matrix(nrow = gridRes,ncol = gridRes)
  for (i in 1:gridRes){
    for (j in 1:gridRes) {
      test4 <- abs((xo[i,j] + sqrt(as.complex(-1))*yo[i,j]) - xy)
      g <- (test4^2) * (log(test4)-1)
      outmat[i,j] <- g %*% weights}}
  outDf <- data.frame(x = xo[,1],outmat)
  names(outDf)[1:length(yo[1,])+1] <- yo[1,]
  return(outDf)}

@@

More recently, Victor and Mast [-@Victor1991] proposed a simpler version of $T^2$, called $T^2_{circ}$. The $T^2_{circ}$ statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance. When these conditions are met, the test statistic for the one-sample case is defined as:

@@

where $x_j$ denotes the $j$th observation of the dependent variables, and all other terms retain their previous meanings. Notice that no covariance term is present in equation \@ref(eq:t2c), because of the independence assumption. This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as we will demonstrate below). Conceptually, this statistic takes the vector difference between the centroid ($\bar{x}$) and a comparison point ($\mu$), and scales by the lengths of the residual vector lines joining each data point to the centroid. Two-sample and repeated measures versions of the $T^2_{circ}$ statistic are also possible.

@@

All scripts used to generate this manuscript are available at: [https://github.com/bakerdh/FourierStats](https://github.com/bakerdh/FourierStats). This includes an _R_ package, featuring functions to implement one-sample Hotelling's [-@Hotelling1931] $T^2$ statistic, Victor & Mast's [-@Victor1991] $T^2_{circ}$ statistic, and the condition index and $ANOVA^2_{circ}$ statistics proposed in this paper.

@@

The upper row of Figure \@ref(fig:amphists) shows scatterplots of simulated Fourier components, expressed using real (_x_) and imaginary (_y_) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \@ref(fig:amphists) shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's _d_ > 2; Cohen's _d_ is the mean difference scaled by the standard deviation, see Cohen [-@Cohen1988]). One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.

@@

Here we replicate and extend these simulations (see Figure \@ref(fig:powerfig)), and confirm that the regime where $T^2_{circ}$ has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \@ref(fig:powerfig)f). However, for effect sizes around 0.5 < _d_ < 1, $T^2_{circ}$ is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N > 32) and large effect sizes (_d_ > 2 when N > 8). These simulations suggest a straightforward heuristic - there is no advantage to using the $T^2_{circ}$ statistic for large sample sizes (when N > 32), so its use should be restricted to small sample studies.

@@

Although $T^2_{circ}$ can be more sensitive than $T^2$, this greater sensitivity relies on meeting the test's more stringent assumptions. The two variables must be independent (i.e. uncorrelated), and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \@ref(fig:falsealarms) shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of $\alpha = 0.05$ (dashed line). However, as the data become increasingly correlated (Figure \@ref(fig:falsealarms)a), or the variances of the two dependent variables more disparate (Figure \@ref(fig:falsealarms)b), the Type I error rate of the $T^2_{circ}$ statistic (shown in red) increases by almost a factor of 2. In contrast, the $T^2$ statistic, which explicitly takes account of the covariance matrix (see equation \@ref(eq:t2eq)) shows no increase (black curves).

```{r falsealarms, fig.width=6, fig.height=9, fig.cap='Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b). Estimates are for 100000 simulations per condition, with N = 10 observations. The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.', echo=FALSE}

@@

  A <- cov(compdata)
  ctr    <- colMeans(compdata) 
  RR     <- chol(A)                               # Cholesky decomposition
  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
  eigVal  <- eigen(A)$values
  eigVec  <- eigen(A)$vectors
  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
  ellRot  <- eigVec %*% t(ellBase) 
  ellRot[1,] <- ellRot[1,] + ctr[1]
  ellRot[2,] <- ellRot[2,] + ctr[2]

  points(0.05*xdata[1:100]+corvals[cond],0.005*ydata[1:100]+0.02,pch=16,cex=0.5,col=nbdtpal[3])
  matlines(0.1*xMat+corvals[cond], 0.01*yMat+0.02, lty=1, lwd=3, col="black")
  lines(0.1*ellRot[1, ]+corvals[cond], 0.01*ellRot[2, ]+0.02, lwd=2)
}
text(-0.95,0.095,'(a)',cex=1.5)

@@

  A <- cov(compdata)
  ctr    <- colMeans(compdata) 
  RR     <- chol(A)                               # Cholesky decomposition
  angles <- seq(0, 2*pi, length.out=200)          # angles for ellipse
  ell    <- 1 * cbind(cos(angles), sin(angles)) %*% RR  # ellipse scaled with factor 1
  ellCtr <- sweep(ell, 2, ctr, "+")               # center ellipse to the data centroid
  eigVal  <- eigen(A)$values
  eigVec  <- eigen(A)$vectors
  eigScl  <- eigVec  %*% diag(sqrt(eigVal))  # scale eigenvectors to length = square-root
  xMat    <- rbind(ctr[1] + eigScl[1, ], ctr[1] - eigScl[1, ])
  yMat    <- rbind(ctr[2] + eigScl[2, ], ctr[2] - eigScl[2, ])
  ellBase <- cbind(sqrt(eigVal[1])*cos(angles), sqrt(eigVal[2])*sin(angles)) 
  ellRot  <- eigVec %*% t(ellBase) 
  ellRot[1,] <- ellRot[1,] + ctr[1]
  ellRot[2,] <- ellRot[2,] + ctr[2]

  points(0.05*xdata[1:100]+corvals[cond],0.005*ydata[1:100]+0.02,pch=16,cex=0.5,col=nbdtpal[3])
  matlines(0.1*xMat+corvals[cond], 0.01*yMat+0.02, lty=1, lwd=3, col="black")
  lines(0.1*ellRot[1, ]+corvals[cond], 0.01*ellRot[2, ]+0.02, lwd=2)
}
text(-0.95,0.095,'(b)',cex=1.5)

@@

plotlims <- c(0,8,0,2) 
ticklocsx <- seq(0,8,1)    # locations of tick marks on x axis
ticklocsy <- seq(0,1,0.2)    # locations of tick marks on y axis
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     # plot tick marks (no labels)
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklocsx, side = 1, at=ticklocsx)     # add the tick labels
mtext(text = ticklocsy, side = 2, at=ticklocsy, line=0.2, las=1) 
title(xlab="Outlier Mahalanobis distance (D)", col.lab=rgb(0,0,0), line=1.3, cex.lab=1.2)
mtext(text = "Proportion significant", side = 2, at=0.5, line=1.5, cex=1.3)
points(0,2,pch=15,col=rgb(0.99,0.99,0.99),cex=0.1)

@@

  polygon(0.4*ellRot2[1, ]+xlocs[m], 0.1*ellRot2[2, ]+1.5, col=addalpha(nbdtpal[2],0.3),border=NA)
  lines(0.4*ellRot[1, ]+xlocs[m], 0.1*ellRot[2, ]+1.5, lwd=1)
  lines(0.4*c(ctr2[1],xdata[1])+xlocs[m],0.1*c(ctr2[2],ydata[1])+1.5,col=nbdtpal[2],lwd=3)
  points(0.4*xdata+xlocs[m],0.1*ydata+1.5,pch=16,cex=0.5,col=nbdtpal[3])
  points(0.4*ctr2[1]+xlocs[m],0.1*ctr2[2]+1.5,pch=16,cex=1,col=nbdtpal[1])
  points(0.4*xdata[1]+xlocs[m],0.1*ydata[1]+1.5,pch=16,cex=1,col=nbdtpal[2])
}

  text(1.25,1.8,'D = 1',cex=1.5,adj=0.5)
  text(3.75,1.8,'D = 3',cex=1.5,adj=0.5)
  text(6.25,1.8,'D = 5',cex=1.5,adj=0.5)

@@

# Controlling for multiple comparisons across location and time

In many studies, it is important to compare responses over space and/or time. However with large numbers of sensors, voxels or temporal epochs, the familywise error rate quickly becomes problematic, inflating the Type I error (false positive) rate. Solutions such as Bonferroni correction, which adjust the alpha level based on the number of comparisons, are overly conservative and can obscure real effects by dramatically reducing power. An alternative approach is to use cluster correction methods to control the Type I error rate, using mass univariate tests [e.g. @Maris2007]. These typically involve summing test statistics such as T-values across adjacent significant locations and/or moments in time. The summed test statistic is compared to a null distribution generated from the same data by randomly permuting condition labels (or the sign of the data for one-sample tests). Such methods control the Type I error rate without substantially reducing statistical power. The same approach can be straightforwardly applied to the $T^2$ and $T^2_{circ}$ statistics. This allows a principled approach to identifying clusters of significant sensors, timepoints or frequencies responding to periodic stimuli. The _FourierStats_ package includes an implementation of this method with options for multivariate statistics (the _clustercorrect_ function).

@@

Following ${ANOVA}^2_{circ}$, one could calculate $T^2_{circ}$ statistics to make post-hoc pairwise comparisons between conditions, providing that standard multiple comparison corrections are applied (e.g. Bonferroni correction). Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary). A repeated measures version of ${ANOVA}^2_{circ}$ has also been implemented following the same logic. The project repository contains _R_ scripts for both of these functions.

@@

@@

load('misc/Hwangdata.RData')

@@

ellRot1 <- getel(data.frame(xdata,ydata))
ellRot2 <- getel(data.frame(xdata2,ydata2))
polygon(10*ellRot1[1, ]+120, 1+ellRot1[2, ]/3, col=addalpha(nbdtpal[2],0.3),border=NA)
polygon(10*ellRot2[1, ]+120, 1+ellRot2[2, ]/3, col=addalpha(nbdtpal[3],0.3),border=NA)

@@

totest <- data.frame(xdata,ydata)
totest2 <- data.frame(xdata2,ydata2)
cisound <- CI.test(totest)
cilight <- CI.test(totest2)
tsqcresult <- tsqc.test(totest,y=totest2,paired=TRUE)

abs1 <- abs(rowMeans(mousedata[,1:2,11,f1]))
abs2 <- abs(rowMeans(mousedata[,1:2,12,f1]))
tresult <- t.test(abs1,abs2,paired=TRUE)

colnames(totest2) <- colnames(totest)
fores <- rbind(totest,totest2)
mahala_sq <- pairwise.mahalanobis(fores, rep(1:2,each=6))
D <- sqrt(mahala_sq$distance[1,2])

@@

Figure \@ref(fig:mousedata)a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz. There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality. Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (_t_ = `r round(tresult$statistic,digits=2)`, _df_ = `r tresult$parameter`, _p_ = `r round(tresult$p.value,digits=2)`). However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \@ref(fig:mousedata)b). The condition index test was non-significant for both conditions (sound: CI = `r round(cisound$CI,digits=2)`, _p_ = `r round(cisound$pval,digits=2)`; light: CI = `r round(cilight$CI,digits=2)`, _p_ = `r round(cilight$pval,digits=2)`), so a paired-samples $T^2_{circ}$ test was conducted. This revealed a significant difference between conditions ($T^2_{circ}$ = `r round(tsqcresult$tsqc,digits=2)`, $F_{(`r tsqcresult$df1`,`r tsqcresult$df2`)}$ = `r round(tsqcresult$Fratio,digits=2)`, _p_ = `r round(tsqcresult$p.value,digits=3)`) with an effect size of D = `r round(D,digits=2)`. This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al. [-@Hwang2019] went on to explore interactions between these two signals.

@@

Vilidaite et al. [-@Vilidaite2018] measured visual responses to flickering grating stimuli in a large sample of 100 adults. Each participant completed a series of 11-second trials, in which stimuli of different contrasts flickered at 7Hz (on-off sinusoidal flicker). Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \@ref(fig:humanSSVEP)), were well-isolated in the Fourier domain, and increased with stimulus contrast. Significant activity was evident at 4% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of \@ref(fig:humanSSVEP). For the main analysis, we took responses from electrode _Oz_ at the occipital pole (black points in the upper row of Figure \@ref(fig:humanSSVEP)), averaged across repetition for each participant. Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \@ref(fig:humanSSVEP). Any participant that contributed at least one outlier was excluded, leaving a total of 89 participants for the main analysis.

```{r humanSSVEP, fig.cap = 'Summary of human SSVEP data. Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes). Electrodes marked in red indicate cluster-corrected significance.  Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point). Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.', fig.width=14, fig.height=5.4, fig.align="center", echo=FALSE}

@@

nexcluded <- NULL
isincluded <- 1 + 0*(1:100)
for (cond in 1:7){
xvals <- Re(subjmeans[,cond])
yvals <- Im(subjmeans[,cond])
bdata <- data.frame(xvals,yvals)
D <- sqrt(mahalanobis(bdata, colMeans(bdata), cov(bdata)))
i <- which(D>3)
nexcluded[cond] <- length(i)
isincluded[i] <- 0}

distmatrix <- matrix(0,nrow=64,ncol=64)
adjacencymatrix <- matrix(0,nrow=64,ncol=64)
for (n in 1:64){
  for (m in 1:64){
    xy1 <- montage$electrodelocs[montage$channelmappings[n],1:2]
    xy2 <- montage$electrodelocs[montage$channelmappings[m],1:2]
    distmatrix[n,m] <- as.numeric(sqrt((xy1[1]-xy2[1])^2 + (xy1[2]-xy2[2])^2))
  }
}
threshold <- 0.18
i1 <- which(distmatrix<=threshold)
i2 <- which(distmatrix==0)
adjacencymatrix[i1] <- 1
adjacencymatrix[i2] <- 0

@@

plot(x=NULL,y=NULL,axes=FALSE,ann=FALSE, xlim=c(0,8), ylim=c(0,2))

for (cond in 1:7){

dataforcc <- apply(alltarget[which(isincluded==1),1:64,cond,],c(1,2),mean)
clusts <- clustercorrect(dataforcc,adjacencymatrix=adjacencymatrix,testtype=3,clustformthresh=0.0001,clustthresh=0.01,nresamples=1000)

datatoplot <- abs(apply(alltarget[which(isincluded==1),1:64,cond,], 2, mean))
datatoplot[c(13,19)] <- mean(datatoplot)   # set the mastoids to 0
testDat<- data.frame(x = montage$electrodelocs[montage$channelmappings,1],
                     y = -montage$electrodelocs[montage$channelmappings,2],
                     z = datatoplot)

@@

for (n in 1:64){points(cond+montage$electrodelocs[n,1]*0.8,1.5+montage$electrodelocs[n,2]*0.8,pch=16,col="grey",cex=0.8)}
points(cond+montage$electrodelocs[blackelectrodes,1]*0.8,1.5+montage$electrodelocs[blackelectrodes,2]*0.8,pch=16,col="black",cex=1.5)

if (clusts$nclusters>0){
for (n in 1:clusts$nclusters){
  toplot <- clusts$clusterpoints[[n]]
  points(cond+montage$electrodelocs[montage$channelmappings[toplot],1]*0.8,1.5+montage$electrodelocs[montage$channelmappings[toplot],2]*0.8,pch=16,col=nbdtpal[2],cex=0.8)}}

@@

allCI <- NULL
allCIp <- NULL
for (cond in 1:7){
  xvals <- Re(cleandata[,cond])/5
  yvals <- Im(cleandata[,cond])/5
  ellRot <- getel(data.frame(xvals,yvals))
  lines(ellRot[1, ]+cond, ellRot[2, ]+0.5)

  text(cond,1,paste(contrastlevels[cond],'%',sep=''),adj=0.5,cex=1.5)
totest <- data.frame(xvals,yvals)
temp <- CI.test(totest)
allCI[cond] <- temp$CI
allCIp[cond] <- temp$pval
}
